\documentclass[10pt,letter,oneside]{scrartcl}
\usepackage{setspace}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
%\usepackage{raleway}
%	\renewcommand*\familydefault{\sfdefault} %% Only if the base font of the document is to be sans serif
\usepackage{graphicx} % Required for including images
\usepackage{enumitem} % Required for manipulating the whitespace between and within lists
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
%\usepackage[round]{natbib}
\usepackage{breakurl}
\usepackage[breaklinks]{hyperref}
\usepackage{endnotes}
\let\footnote=\endnote

\usepackage{authblk}
\author[1]{Luis Felipe Rosado Murillo}
\author[2]{Christopher Kelty}
\affil[1]{Berkman Center for Internet and Society, Harvard}
\affil[2]{Institute for Society and Genetics, Department of Anthropology, and Department of Information Studies, UCLA}
\renewcommand\Affilfont{\itshape\small}

\title{Digital Expertise as a Cultural System}

% I changed from "computer collectives" to "collectives of digital experts" for two reasons: 1) broaden the scope, since "computer" already sounds too much like "typerwriter collectives" and 2) because "anthropology of expertise" is a known thing across the field.



\date{}

\begin{document}
\maketitle
\section{Abstract}
In this article, we offer some well-established words of advice on
ethnographic method and some new insights into the specificities of
collectives of digital experts. New forms of digital expertise call us
to rethink ethnographic research design and practice. We discuss
issues of building rapport, creating empathy, dealing with different
expectations, and conflicts of interpretation with examples from our
field research and suggestions for advancing new forms of
collaboration through the study of emergent forms of technical and
political participation in the domain of computing technology.  As a
concluding argument, we elaborate on the importance of the practice
and the study of collaboration in the context of science and
technology studies more broadly.

\doublespacing
\section{Introduction}

What difference does it make to fieldwork in cultural anthropology to
study collectives of digital experts?  How different is this domain
from ``traditional'' subjects or other domains of contemporary
anthropology?  In this article, we offer some well-established words
of advice on method, but also some new insights into the specificities
of collectives of digital experts that call us to rethink ethnographic
research design and practice.  Our interest is not in the use of
digital technologies---which is necessarily of great interest to all
researchers today---but in the norms and forms of life of those who
invent, hack, reconfigure, build and distribute such technologies. 

``Collectives of digital experts'' is our term of art.  It is intended
to signal all those sets of people who consider themselves technical
experts in some broad sense: professional engineers, software
developers, hackers, makers, amateur scientists and technologists from
many technoscientific domains.  It is also intended to signal that
these experts form new kinds of collectives (as well as joining in
older forms), and that there is a relationship between the expertise
and the collective.  Doing fieldwork amongst such groups demands
different research practices than doing fieldwork on \emph{users};
even if the putative boundary between experts and users is blurry, it
is nonetheless constantly reinstated and renegotiated.  This
difference in focus is also meant to hold digital technology as
relational \emph{substance} more than as \emph{media}.  There are
already different schools and traditions of emphasizing the
relationship of the cultural, the material, the political, and the
digital, and we try here to distinguish them.

Key questions in any kind of ethnographic research are questions of
\emph{rapport}, \emph{empathy} and \emph{complicity}.  These are not
merely methodological questions of access but concern the cultivation
of ethical relationships, the reliance upon and sometimes competition
with multiple constituents who may or may not align with the goals of
a dialogic ethnographic inquiry. Relationships in the field are built
from sustained and often intense interactions that create binding ties
and are characteristic of anthropological field research. The course
of a project depends on these relationships, if it is not entirely
determined by them.  What one makes of these relationships, and the
insights gained thereby constitutes the core of any ethnography worthy
of the title---but the nature of these relationships is obviously
different in different places: what is true of interactions with the
Gawa of New Guinea is not necessarily true of interactions with the
Bororo of the Amazon, much less of those with young engineers in a
makerspace in Shenzhen, Tokyo, or New York.

Before building ties, the question of ``entering the
field''---inadequately summed up in the term ``access''---presents
itself as the first challenge.  Modes of presentation, perceptions and
usages of differences of cultivation, self-identification, and
belonging along socioeconomic, political, ethnic, and gender lines,
veiled and overt asymmetries in relations of power---all these things
account for multiple determinations of the ethnographic encounter. One
can be invisible or too visible, the object of constant attention
and/or suspicion at a shared workshop, computer lab, or online space,
or operate anonymously or pseudonymously by nickname or avatar image.
Regardless of the circumstances of a first encounter, one cannot
``just show up'' to do fieldwork.  When it comes either to the study
of sociotechnical phenomena in established or marginal communities of
computer technologists, there are specific issues at stake, which take
time and effort to understand.
  
Questions of method tend to vary greatly by object and
(sub)discipline--- medical anthropology has its own standards for
interaction with and the development of relationships with patients,
doctors, and other caregivers or healers, just as ethnographers
working in the domain of psychological anthropology do when developing
particular techniques of person-centered ethnography \cite{Hollan1992}
and the study of emotional states and expressions
\cite{Throop2009}. But for most anthropologists, ethnography is the
result of entanglements in social and cultural life that are engaged
in because they reveal something of ourselves and others, which cannot
be easily articulated. Those entanglements can rely on involvement and
complicity, or on the conscious adoption of ``adjacency''
\cite{Rabinow-MarkingTime}, but not on distance, critical or
otherwise, that takes social life to be observable without engagement
in it.  In the case of new collectives of digital experts, a different
relationship towards these long-standing anthropological modes of
encounter may be necessary, given the specific horizon of the
contemporary within which such collectives exist \cite{Rabinow2008}.

In this article we will discuss these issues with examples from our
field research and suggestions for advancing new forms of
collaboration through the study of emergent forms of technical and
political participation in the domain of computing.  When rethinking
ethnographic practice in the context of computer expert collectives,
we draw from the contemporary developments in the domain of
multi-sited ethnography
(\cite{Fischer1999,Marcus1995,Burrell2009,Falzonetal2009,Rabinow2008})
and ethnographic studies of virtual worlds
(\cite{BOELLSTORFF2008,Miller2001,HorstandMiller2012,BoestorffNardi2013}),
as well as established ethnomethodological frames in STS
\cite{Forsythe1999,forsythe2001studying,Latoour2007,Suchman1987,Suchman1999}.  As a concluding argument, we
propose the importance of the practice and the study of collaboration
and ethnographic composition \cite{Kelty2008a} in the context of
science and technology studies more broadly.

***

In previous work we have articulated ethnographic fieldwork as, in
part, an ``epistemological encounter'' \cite{Kelty2008} in which the
experience of participating in the social, ethical and cultural lives
of others necessarily forces one to revise one's own theories and
beliefs.  Those theories and beliefs can be unstated, as part of one's
own culture (``making the familiar strange'' as the canonical formula
goes), or imported as part of training and intellectual discipline
(``theory-laden observations'').  If fieldwork experience does not
produce an encounter of this sort, then it can have little more than a
confirmatory or extremely weak (vis-á-vis other methods)
hypothesis-testing function, not one of discovery or conceptual
refinement or invention.  More recently in anthropology, a similar
claim has been made about the need for an ``ontological turn'' with
radical conceptual work by which anthropologists encounter different
modes of being \cite{VdC,Latour,Kohn,Holmrad}, especially those that
are radically different from the philosophical and theoretical
traditions in which most anthropologists are trained.  There is
clearly a relation to the claim of an ``epistemological encounter''
---but we do not propose to resolve this relationship here.
Nonetheless, both approaches would demand that we treat carefully the
specificities of the particular subject of study---which in our case
means collectives of digital experts.

In what follows we will discuss issues of exchange, rapport, conflict,
and collaboration for the ethnographic study of sociotechnical
phenomena in the domain of digital computing. We will also describe
practical challenges for sustained participation in order to emphasize
the transformative potential of collaborative theorizing for
anthropology more broadly.

\subsection*{Ethnicity, Gender and Other Power Dynamics}

Ethnic distinctions, gender roles and expectations, and variation in
cultural beliefs and moralities have historically constituted the very
core of any ethnographic project. Being perceived and identified in
the field under a certain form of classification overdetermines the
course of interaction and, ultimately, the course of ethnographic
research, rendering in symbols and meanings sociocultural processes of
broader scales. Anthropologists have a rich tradition of
understanding, critique and methodological innovation with respect to
such overdeterminations. 

In our fieldwork experience amongst digital expert collectives,
examples abound, specifically those that are related to invisibility
and hypervisibility of perceived and marked differences.  Far from
being the utopian space in which ``no one knows you are a dog'',
conventional forms of identification saturate the lives of digital
experts.  

While ethnicity plays a less obvious role than gender in collectives
of digital experts, it is nonetheless clearly marked out.  At the
level of international global circulation of technical experts,
certain ethnicities and nationalities invoke immediate judgements or
expectations, as has been the case for instance of migrating Indian software
experts \cite{xiang2007}, or reflect more local concerns with ethnic
identification \cite{Takhteyev2012,Burrell,Chan}.  Reconfigurations of
ethnicity in user-space are arguably much better studied and
understood through the lens of anthropological studies of personhood
and race \cite{BOELLSTORFF2008,Miller,Nakamura}.

In respect to gender identifications and roles such issues
express themselves as forms of competitive comparison, exclusion
through veiled silencing, and association of gendered identities with
particular political forms and technical capabilities.  For example,
in one case a prestigious engineer and hackerspace founder, when
comparing herself to another prestigious woman engineer declared,
``she has a bigger penis than mine,'' largely in order to indicate and
compare levels of technical knowledge or prowess.  The language of
comparison is often playful (cf. \cite{Coleman2012} on joy) and, at
times, harmful, and may include both self-reflexive markers of gender,
class, race and religious differentiation as well as unrecognized and
unstated assumptions built into the very commonplace joking and
jesting in person and on-line.  Many marked differences related to
gender, for instance, are established and re-negotiated through
established forms of technical engagement, and resolved through
technical arguments and performance of computing knowledge.

For instance, there are many masculine expressions of independence,
bravery, and, ultimately, power: ``real men do not use debuggers,'' as
the tongue-in-cheek expression goes, or the nostalgia for a ``time when
men were men and wrote their own device drivers'' once expressed by
Linus Torvalds, an influential and controversial Free Software
developer, in the context of the Linux kernel development
mailing-list.  Such expressions disguise other voices: they express
both an anxiety about masculinity traditionally denied to brain-work
and ``nerdy'' occupations \cite{BenjaminNugent}, at the same time that
they re-establish gender distinctions around being ``close to the
machine'' \cite{EllenUllman} or technically proficient at ever more
arcane and often useless pursuits.  Such assertions are very much like
the classic portrait that Veblen drew, where ``device drivers'' and
``having code accepted in the kernel'' or ``having a Open hardware
project that is widely used'' are ``trophies'' that signal invidious
distinctions with respect to the value of labor.  By contrast, what is
considered menial work--- documentation, code-testing, stocking the
hackerspace and cleaning it---become ``drudgery'' all too often
associated with the feminine \cite{Veblen-Leisure}.

The fieldworker who enters without such skills therefore, is quite
likely to experience these distinctions from below---as a newbie or
lurker---in some spaces more than in others.  There are plenty of
hacker and maker spaces that have emerged in the last five years which
are explicitly focused on rethinking the practices of inclusion and
exclusion to re-evaluate and revalidate the collaborative labor
involved---but not all sites are the same. One of the important
aspects which determine the attitude toward inclusionary/exclusionary
practices has to do with the trajectory from which the original group
of founders is assembled. Thus, members of corporate sponsored
hackerspaces or the more assertively masculine ``Tech shops'' are
often much less open to discussion and deliberation about social
dynamics and forms of symbolic violence than are members with
experience in autonomist organizations and social movements, even when
those problems are not solvable in the span of months of
consensus-based attempts at problem-solving.

Digital experts are also quite often positioned in higher
socioeconomic positions than the ethnographer. They tend not only to
possess more economic capital, but more prestige in the gendered order
of science and engineering. This difference echos an observation once
made by Durkheim at the turn of the past century, a period of
institutionalization of sociology as an academic discipline in France,
that it was quite obvious that disciplines that served industrial
societies would be held with much more prestige and institutional
support than the human sciences \cite{Durkheim??}. Fast forwarding to
the contemporary, we witness computer scientists engaging in
discussions of culture and computer-human interactions more often,
aiming for a position of power in the ordering of academic disciplines
which extends to engineers and technologists outside academia. Having
in mind this power imbalance is important when working in the field to
create the conditions for engagement.  Without such an awareness, one
is confined either to a certain segment of a social hierarchy
(e.g. newbies and other de-valued forms of work), or condemned to
exercises of ``drive-by ethnography'' with rather busy, and
hard-to-reach elite technologists.  Understanding such forms of
cultural distinction is essential to achieving rapport with the people
who seeks to understand (whether that be those lower down or higher up
the hierarchy).  

Working mostly online can help when most of the work in a certain
anthropological problem space tends to be office work, but it cannot
dispense with regular field visits to professional conferences and
informal gatherings for computer professionals, which might include
more permanent socialization spaces such as cybercafes, hackerspaces,
and maker spaces or transient but no less important events such
documentation fests and collective hacking sessions held for Free and
Open Source development projects
\cite{ColemanAQ,BurrellGhana}. Ultimately, given power imbalances, it
is important to observe, as the anthropological \emph{sensus communis}
dictates, one cannot just ``show up'' but rather, must find ways to
enter webs of relationship (which involve computer technologies and
technologists equally) of those who are already ``in'' so to speak.

One important difference when studying collectives of digital experts
is that academic titles and institutional affiliations do not
influence or facilitate the process of getting in, except for a few
cases of recognized research centers (mythical places and ```places of
memory'' like the Massachusetts Institute of Technology) given their importance to the history of
computing. Such affiliations are not reliable diacritical elements of
distinction and trust, and they often backfire when attempted.  This
is consistent with an ethic of public demonstrations of expertise as more
significant records for attaining prestige, than titles and
affiliations. In certain contexts, academic affiliations are highly
respected but these contexts also involve the positioning of the
ethnographer in the reversed imbalance: that of an origin of
socioeconomic and political power over the location of origin for
research co-participants.

\subsection*{``When you Study us, We Study you Back''} 

In our field experiences and the experiences of colleagues we have
learned that there is always a great potential for misrecognition on
many sides of the ethnographic equation. We have quite often come
across researchers who approached underground computer groups with
objectifying motives which resulted in distancing themselves from the
group and foreclosing any possibility of building rapport.  People
presenting themselves as interested researchers on mailing-lists, for
instance, quite often fail when trying to all-too-easily gather data.
Especially when dealing with self-reflexive, academically minded,
critical experts (whether hackers, lawyers, scientists, doctors),
asking for interviews or surveys questionnaires to be filled out can
be a mine-field.  Awareness of the ``metapragmatic'' conditions of the
research interview or interaction is essential to for making sense of
what is communicated in any given setting \cite{BriggsLearning}.  The
example below illustrates one of these cases which deserves full
consideration since it is common in the experience of hackerspaces.

\begin{quote}
  A graduate student registered to the public mailing-list of a
  hackerspace and sent a request for members to fill out an online
  survey. In his or her self-presentation, the student presented in a
  few lines the research objectives which treated the ``resurgence of
  the Maker movement.'' The questions appeared to be market-research
  oriented --- with inquiries about electronics purchase habits, etc.  
  A hackerspace member replied, stating that the research was
  flawed. Another member then responded saying that he was offended by
  the fact that the researcher seems to have missed the main goal of
  the hackerspace, which is that rather than consuming electronics, they
  are all invested in creating their own. Turns out the researcher's project 
  was for a course on ``consumer culture,'' which sparked strong reactions. 
  ``It's pretty offensive (to some of us at least),'' a founding member 
  sentenced, ``to be `studied' in the context of a `consumer culture' class''. 
  Finally,  another member joined the exchange, after investigating the list of 
  classes one has to take in order to get a masters' degree in ``Creative Brand 
  Management,'' and pinpoints it: ``welcome to {[}the hackerspace{]}, where 
  your research subjects may research you back.''\footnote{Source:
    https://www.noisebridge.net/pipermail/noisebridge-discuss/2010-July/015115.html}
\end{quote}

This example calls attention to the fact that computer cultures
themselves have a strong component of academic research culture. They
are not only invested in academic research or academic topics of
research, but dedicated to the process of cultivating by themselves and 
with help of mentors and friends the technical skills they learn to desire
in order to become the technologist they aspire to be. 

There is also here a clear element of competitiveness (often
gendered--see below), whereby the public performance of this
willingness to ``study you back'' signals the possibility for (indeed
demand for) an exchange of ideas and arguments that might satisfy
those involved in some form---in this it mirrors the experience of
some who study theological debate through ethnography and are often
called upon to enter into such debates
\cite{FischerAbedi,Susanharding,Tuhami,HirschkindDawa}.  Being
accepted can sometimes be contingent on a willingness to defend one's
ideas amongst potential informants, not just amongst academic peers.

\subsection*{Questioning Legitimacy, or Who Gets to be the Expert?}

Competitiveness and collaboration are key facets of work amongst
computer technologists for whom the question of who gets to become a
legitimate actor emerges in spaces in which expertise is highly
distributed and academic accreditation is not considered a substitute
for embodied and demonstrable, public knowledge.

For Kelty \cite{Kelty2008}, this issue was painfully explicit in the
case of Free Software because long before arriving at the topic,
hackers had themselves arrived at a rich, native self-description, and
had their own self-appointed anthropological ``Big Man,'' Eric Raymond.
To enter this field as a trained anthropologist, but lacking the
experience and participatory centrality meant that Kelty was at a
disadvantage in proposing alternative explanations for what was
happening in this community.  In this case a particular
``non-encounter'' was the result---for the native anthropologist, the
audience was composed of hackers and a playful enjoyment in making
themselves strange---and not academic anthropologists who were
generally not engaged at all.  By contrast, the salient audience for
Kelty was both hackers and academic anthropologists, and the challenge
was to find a way to reflect the experience of participation in both
directions, through, for instance, the re-telling of the history of
Free Software (which appeals to many who lived through it) and the
refinement of the concept of ``recursive publics'' which generally had
more appeal to academics than hackers.

In many cases, the question of ``who gets to speak'' can lead to heavy
criticism and even exclusion from public channels of communication.
In one occasion, for example, one colleague of ours was invited to
give a talk at an foreign university. The announcement of her talk was
circulated widely among university students and it ended up in one of
the oldest and the most active Linux Users groups in the
region. Unexpectedly, a ``flamefest'' ensured---in this case not among
members of the list, but directed against the anthropologist who dared
to speak of Free Software. Using disproportionate anger and strong
language, one of the mailing-list members said ``Hacker culture is a
hyped subject, it is at the tip of the tongue of many people who are
engaged with Free Software, however, despite much discussion, no code
will be generated. Unfortunately, the hardware we use depends on Free
Software, not philosophy, to work {[}...{]}'' (\emph{our
  translation}). Another member of the list then replied: ``cat
filosofia \textgreater{} /dev/null'' in a typical code-switching
manner that characterizes much of online exchange among computer
experts and hackers, expressing the need for discarding philosophical
discussions.  The potential relations of this particular
anthropologist with the local community was henceforth shut.

  % Can also use the example here of "shut up and show me the code" in
  % two bits.  

As this case illustrates, legitimacy tends to follow from a technical
work ethics and cultural sensibilities: hard work is measured in
technical contribution, style, and in public displays of
expertise. Building rapport is conditioned upon a level of technical
expertise coupled with the capacity to draw boundaries between what
one is doing (research) and the task at hand. It can vary considerably
in scope to involve one or a few developers, or a transnational
collaboration across languages, spanning continents and bring multiple
experts into collaboration.  Such tasks might include developing a
system library, debugging a digital circuit, collaborating on the
heavy lifting task of debugging, documenting, packaging, and
distributing software for an operating system, testing and documenting
a new prototyping platform, and so on. Thus, one of the stronger
values we find in the context of digital or computing collectives is the value of
work: it has to be taken under careful consideration (as a question of
theoretical interest) as much as a means for actual, fruitful
participation (the important and downplayed side of ``participant
observation'').

From a pedagogical standpoint, these facts also suggest an answer to a
question frequently asked by novice anthropologists eager to study
digital experts: ``How much technical skill do I need to participate
in these communities?''  The answer is, as much as is necessary to
build exchange relationships with the people one wants to study.  What
has traditionally, in anthropology, been understood as an issue of
linguistic and cultural competence amongst a given population, is
extended here to include a level of technical, engineering or
scientific knowledge (or enthusiasm, at least), which can be brokered
into a set of relationships with similar enthusiasts. One must accept
and return the technical gifts that are given, or to put in other
terms, one has to find one's position in the web of relationships, so
as to give, accept, and re-tribute in different ways as the group
demands. These include not solely technical contributions, but actual
community organizing work---a very valuable and often downplayed
aspect of, for instance, community centers for alternative computing.
Legitimacy and trust flow from these activities, which are in fact
ways of becoming a digital expert---not just studying them.

It is important to bear in mind that collectives of digital
experts---with important exceptions---tend to draw, institute, and
monitor the symbolic borders between technology and society, politics,
and culture. Thus, it is important to avoid being identified solely as
an expert in \emph{some other domain} when working with digital or
computer collectives, as rapport depends on actually working alongside
computer technologists on computer technologies.  In other words, to
be identified only as a philosopher or social scientist who has no interest
in the technical aspects of the hacker lifeworld tends to be quite
detrimental for the actual progress of the research.

Where STS scholars see inseparable realms of sociotechnical activity,
technologists themselves quite actively work to draw distinctions and
tease apart, so as to denounce, efface or negate the sociocultural and
historical foundations of their practices.  This is as much an object
of inquiry for us as a practical dilemma and a contentious topic
regarding our participation and collaboration with computer
collectives. It speaks to the ``worth'' of our work \cite{Stark} and
our presence (ultimately our ``worth'' as persons) which responds to
criteria that is foreign to the legitimization, validation, and
recognition criteria we are subjected in academia.  It consists in a
double-bind involving the research practice, the need for advancing
empirical studies of computing and theorizing of sociotechnical worlds
through active participation and collaboration.

The nature of this kind of rapport amongst digital experts also raises
an important question about the kinds of collectivities they form.  To
do anthropological research in a traditionally organized firm---where
bureaucracy and hierarchy govern many of the necessary forms of
exchange and transaction---requires negotiating that collective or
organizational structure (is one simply \emph{hired?} Can one ``work''
alongside them unpaid? Or should one attempt to enter the lifeworld of
the firm by different routes?).  By contrast many emergent collectives
of digital experts are constituted outside of firms, through other
networks and collective arrangements---often (but not always) more
open to curious onlookers.  Such differences both afford participation
and restrict it in ways that need to be made explicit.

Participating in hackerspaces we came to learn about one of their core
legitimacy criteria: in order to be regularly present and become a
\emph{bona fide} member, one has to engage in collaborative efforts to
maintain the space and advance projects---helping others and advancing
their technical projects, as well as one's own projects.  A case in point for Murillo was
the development of a large sensor network across Japan, used to
measure levels of radioactive contamination after the Fukushima
disaster.  The project required testing and using digital circuits,
prototyping boards and projects for certain purposes, or
self-education with an openness to sharing what one has learned with
others, to repurposing technological artifacts for the purpose at
hand.  Such deep involvement in work and building, knowledge
production, sharing and critique, are at the heart of what it means to
become expert in such collectives, and the anthropologist is naturally
confronted with the same challenge.


% \subsection*{Collaborative Theorizing} 

%% CK -->  I think this is a different paper, actually, and could be
%% excised from this one and maybe be the basis for another one...

% The potential for collaboration in academic research leads us to a 
% different and often productive aspect of work amongst certain kinds 
% of technical experts: the capacity for collaborative theorizing.  
% This can be accomplished in many ways, but we have found the 
% conceptual work around ``abstractions'' in the domain of computing to
% be particularly useful. Identified as a foundational concept for 
% managing complexity in computer systems or data structures, it 
% also creates expectations and demands on the part 
% of those who engage in the practice and knowledge production of 
% computing. It is around the notion of complexity, for instance,
% that forms of interoperability across systems (and levels within
% a system) are built. Not only when dealing with technical artifacts,
% rich abstractions are used also in the context of everyday interactions
% at hackerspaces and other technical places of exchange. Metaphors from
% the domain of computing are often used to talk about community 
% organization (i.e. in the context of the hackerspace network, ``design
% patterns'' were created and have been revised to define broad guidelines
% for starting, managing, and running community spaces---under the 
% direct influence of software development design patterns).
    
% % I think this should be an ontological turn point.  It is not
% % about exchange with different disciplines, it is about the uses to
% % which alterity is put.  So free software is not a discipline, but
% % for many a very alien mode.  As are many such technical subcultures.
% % not as alien as Amerindian multnaturalism, but at least as hard to
% % understand.
% %
% % LF: more soon!
   
% Forms of exchange are key points of entry for anthropological reflection,
% theorizing, and dialogic engagement, in particular, with respect to the
% contemporary anthropological literature on gift exchange (Strathern 1988;
% Weiner 199x; Graeber 2001; Caillé 2000). It is important not only to 
% explore dynamics of technical exchange (involving the circulation of 
% software, hardware, documentation or technical help for instance) but 
% to actually illuminate how resources are shared in order to participate 
% and collaborate as an ethnographer. Based on different classes of 
% cultural goods, their symbols and extensions of particular technical 
% persons or groups, we find particular unparalleled opportunities to 
% make sense and participate in our fieldsites. Put differently, what is
% already known in the literature regarding webs of reciprocity and
% interdependence can serve as entry points for meaningful participation.


\subsection*{Research caught in accelerated  temporalities} 

Marx famously argued that social life was being determined by the temporality 
of the machinery and of the productive forces (in Capital):
the metaphor of our times is not that of the \emph{need,} as an
objective economic force, an injuction to keep the ``gears'' of
value-extraction and accumulation turning, but to keep ``data
gathering and flowing'' for similar---or allegedly subversive---purposes. 
Marx's discussion of the rhythm of work as subordinated ``to the 
mechanical needs of the machine itself'' \cite{Marx1990} (p.34) can serve 
as a foil for an important transformation in our notion of time with
the ``computerization of production and circulation.'' 

It is not a mere substitution of the mechanical for the digital and
computational, but rather an observation that structurally and
cosmologically the fundamentals of capital accumulation are preserved
and transposed onto new digital enterprises.  In terms of our research
practice with digital experts, we encounter seemingly insurmountable
differences in respect to time compression. Cycles of work, funding
and of the excitement and enthusiasm of technological fads, new
techniques and tools, or new sources of capital, circulate much more
rapidly than the temporality of academic observation and production.

As a consequence, our time---in the human sciences---has also been
accelerated, but unlike the rhythm of work in computing, we
have a different temporality of knowledge production: our research
takes much longer to conduct, our reports take more time to prepare,
and our publications take much longer to see the light of day.  But
there are related reasons---indeed, shared dispositifs---which
underlay our own practices as well: the precarious nature of our
editorial workflows, based on volunteer help, adjunct and temporary
labor, and overworked faculty members; corporate publishing platform
lock-ins, and intellectual property constraints.  Thus differing
temporalities pose both problem and opportunity when it comes to
studying collectives of digital experts.

The simultaneity of events and forms of interaction on and
off line creates an overwhelming sense of ``ethnographic data
deluge.''  This can have the potential to be quantitatively much superior than in
conventional ethnographic document collection (given constraints of
production, analysis, and transportation of such materials by a sole
ethnographer), but it can also be temporally and emotionally
overwhelming, as it produces the pronounced effect of culture changing
faster than it can even be recorded.  This however, is nothing new for
cultural anthropologists, and is often as subject of debate--from the
salvage ethnography of the early 20th century, to Bateson's
experiments in capturing social life in \emph{Naven}
\cite{Naven,MarcusOnNaven} to the debates about Anthropology as
cultural critique \cite{MarcusFischer,WritingCulture}.

The sense of being overwhelmed by data,is related to a different
aspect of temporality that also frequently afflicts researchers.  This
is the ability to locate the ``object'' of research within the rapidly
changing norms and forms of life in high tech, hacker, and
entrepreneurial worlds of IT. To identify the underlying assemblages
or dispositifs \cite{Rabinow2003} which account for the surface
temporality of things like new devices or applications, new fads and
funding priorities which always participate in a rhetoric of radical
novelty.  Hidden within the accelerated temporalities are other
times---times of maintenance or breakdown, as well as longer term
rationalities related to the structure of an industry or regulatory
environment or social movement within which experts operate day to
day.

From one year to another, companies, technologies and platforms can
rise and fall, and so pegging one's work to a particular
technology---an ethnography of SnapChat, for instance---can only bear
fruit if the ethnographer works at the pace of Silicon Valley and its
pathological economic and practical demands.  But to identify other,
more widely shared ``objects''---often not self-identified by hackers
or other technologists, but attributed by the anthropologist---is
ultimately more true to the logic of discovery in anthropology.

It is a simple point, but the salient feature of a web service such as
Snapchat for an anthropologist is not, or should not be, the app or
the user experience of the app (for which SnapChat should by rights be
paying its own employees to do research), but the ethical, legal and
practical dynamics of forgetting and data permanance, which might be
got at both through those using (and hacking) the app, understanding
the economic and regulatory environment, as well as those creating it
or its competitors and descendants.  Being able to articulate a domain
of questioning remains the most urgent task for anthropologists--which
is not identical with ``finding a site.''  The perceived
``acceleration'' of technology in these worlds is often in stark
contrast to the oft-repeated and unresolved ethical and cultural
structures that re-emerge and recur---and which the anthropologist can
make visible, but the technologist trapped in his or her own
temporality often cannot.

						  %  Though I would say that this is also the
                           %  reason to find a really good principle
                           %  informant-- becuase he or she is the one
                           %  who can do exactly this work and is
                           %  often more than just an expert... but
                           %  hard to find. even harder to befriend. 

Simultaneity represents the potential for a rich multi-sited
research project, but also a huge challenge for ethnographers working
solo, which calls upon the need for creating collaborative research
projects with fair distribution of the work load with due recognition.
The accelerated temporality of fieldwork and the imposition of
circulation across many sites (with all the difficulties that it
represents in terms of time and resources, not to mention all the
legal restrictions for circulation of researchers from the Global
South) points to the especially pronounced difficulties of multi-sited
research on global collectives of digital experts.

A similar issue has been identified with the perils of ``ethnography
by appointment'' described by Hannerz \cite{Hannerz-Appt}. The
transnational condition \cite{Ribeiro1994} under which most research
with computer technologists is conducted may be marked by encounters
with busy professionals at coffee shops and surrounding their work
places. In these cases, it is important to strategize in order to find
ways to \emph{get involved} as a condition of doing actual fieldwork,
instead of a semi-structured interview and web page-scraping work
taking place under the increasingly questionable usage of the term
``ethnographic research.'' This problem speaks about the necessity of
becoming entangled with and participating in technical work, political
activism and other aspects of the hacker lifeworld.


\subsection*{Hablamos C++?}

The importance of speaking ``native'' language has been a key
expectation of ethnographic work since Malinowski---even if it is
often honored in the breach (as it was by Malinowski himself, who
relied on translators, even though he insisted that 3/4 of the 
success of an ethnographic project depended on the ability to 
speak the native language \cite{Young199x-Biography}).
 
The very notion of language, however, as in ``what counts as
language'' and ``what practical usages does it have'' has been the
object of various analytic enterprises in the domain of language
studies so as to encompass interactional, ideational, indexical,
historical and political dimensions. Yet, there is a vast uncharted 
territory to be investigated with respect to the contemporary 
intersections of so-called natural languages (what we would 
deem sociocultural and historical) and artificial languages 
(being those instruments to interface humans and computer machines
with translation of languages with levels of discontinuity that are
untransposable). It is in this sense of the mastering of languages 
broadly construed that the collaborative work with computer expert 
groups depends upon -- broadly construed to refer not solely to 
computation but to other levels of abstraction as well where 
we find digital experts at work concerning sociocultural processes
and sociotechnical entanglements: computing serving as infrastructure 
and culture at once: operating systems and social systems overlapping. 

Methodologically, it is important not to simply be a computer
scientist when studying computer scientists or a hacker when studying
hackers, which means entering the field without any cultivated
certainty that one intimately knows what the field is all
about---essentially erasing the possibility of ``epistemological
encounter.''  Rather, it is fundamental to have the energy and time
commitment to learn new languages of particular worlds (in the
phenomenological sense) as they present themselves to the
ethnographer.  As it rings true to accumulated experience in the realm
of anthropology, one can anticipate questions but not what one will
find in the field. The capacity to deal with the unexpected cannot be
systematically taught in graduate school, it must also be exercised in
the field; which is why most programs teach ethnography through
reading of classic and contemporary ethnographies as well as through
workshops for ethnographic writing.

In the context of computing, familiarity with certain ``field
languages''---another debated keyword of this tradition---is useful from the
outset so one is not left out of conversations, debates, and
activities, depending on the group, traversing often over a broad range 
of technical topics, such as intellectual property, programming languages, 
mathematical theorems, political theories and traditions, theoretical 
computer science and AI, amateur and software-defined radio; as well as
contentious topics which reveal themselves as opportunities to display
one's expertise and worth or to draw distinctions among group members
regarding debates about pros and cons of popular or obscure
programming languages, computer architectures, operating systems,
approaches to software and hardware development, text editors of
choice, etc. Familiarity, therefore, allows for further engagement and
it is usually a good idea to invest in getting accustomed to these
languages before starting the project. Even though, it is quite
certain as far as the anthropological \emph{sensus communis} goes,
that one has to be open to the surprise of the unexpected.


%you are here

\subsection*{Staring at Computer Screens, Holding Soldering Irons}

Ethnographic studies of computing are situated in between faces and
screens, between purportedly distinctive realms of activity \emph{as
  one}---even if one attends to the sociotechnical processes that
underlie, mediate, and undercut differing forms of sociability on and
off-line. Earlier research work has been grounded in the premise that
sociocultural processes, forms of belonging, and the very conditions
of fieldwork research would substantially change with the increased
usage of online communication as to study the online experience in its
own right \cite{Turkle1984,Hine2000,Boelstorff2006} or challenge the
assumption that a new sphere of sociability demanded reformulating our
understanding of field research \cite{MillerandSlater2000,Malaby2008}.

This form of dualism is may longer be sustainable or, for our
purposes, even useful for fieldwork research among collectives of
digital experts: it fails invariably to capture what are the richest
phenomena of all, that of the technical and institutional conditions
for creating virtual worlds on the one hand, and on the other, the
feedback mechanism that the virtual sociability has over interpersonal
interactions both online and off.  There are important
exceptions---it is just as difficult to imagine a fieldwork situation
today that \emph{does not} have to deal with some aspect of digital
communication or interaction as it is easy to imagine situations where
only processes have to be attended to primarily online (mailing-list
discussions, IRC conversations, blog posts, Twitter feeds, bug
trackers, ticket systems, and source code management systems,
etc.). \footnote{See: \cite{BoelstorffandNardi2013} concise handbook on
  ethnographic studies of virtual worlds in the ironic format of the
  ``Notes and Queries in Anthropology'' which had many versions since
  the early development of British Anthropology and was meant to
  prepare ethnographers to do fieldwork anticipating questions of
  theory and method. One of our senior mentors who did fieldwork in
  the 50's among the Inuit says that it was the only book he brought
  to the field and it was completely useless (personal
  communication).}

We would prefer to suggest that it is the blurry line between users
and experts, as opposed to that between real life and virtual reality,
which accounts for many of the most troubling and significant impacts
of the digital.  Much of the discourse about free software, hacker
spaces, anonymous, citizen science, participatory culture and so on,
has insisted that the distinction has been dissolved---users are
experts now.  But this erases distinctions
rather than proliferating them: what is at stake is a reconfiguration
of the relation amongst users of technologies and their ability to
become proficient in their use, institutions that coordinate such
learning, and the legitimacy of that knowledge.

It is clear that not all collectives of digital experts are equally
open or accessible to everyone (including those who would study them),
but there has been a transition by which whole new collectivities with
considerable power have come into existence through forms of
enculturation, enskillment and entrepreneurial self-making and that
seek both to achieve specific political and technical goals, and
(perhaps not always) to reproduce themselves as a collective.  Free
Software, hackerspace collectives, and anonymous are such new
collectives---they are more than social movements but less than
``organizations'' and we do not yet have a solid understanding of how
they are changing the relationship between knowledge and power in
society today. Such collectives of digital expertise are are, at
times, the co-authors of, and have great affinity and responsibility
for, the technologies by which these new forms of human interaction,
exchange, and digital inscription take place and afford change. This
specificity is important to emphasize due to its overdetermination of
field research among collectives of digital experts: technology is not only a
medium (for political speech for instance) but also substance (for
self-fashioning and for the fashioning of alternatives, prefigurations
\cite{Graeber-prefig} or ``concrete utopias'' for example \cite{bloch-hope,Broca}. 


% commented the following paragraph out because I think it opens up
% new questions, rather than wrapping up the section above... 

% In order to cope with the accelerated temporality of field research and
% intensified simultaneity of sociocultural processes across virtual and
% actual domains, field researchers have to devise strategies to work
% under emergent conditions and new expectations of participation and
% collaboration: to approach documents in the field as what in
% anthropology are called ``informants,'' that is, individuals with
% cultural intimacy and native competence who are engaged with the
% ethnographer at the task of interpreting their cultural contexts. Kelty
% and Landecker (200x) discussed this approach in terms of approaching a
% particular piece of publication in context as an anthropologist would
% approach an informant to share the task of interpretation of cultural
% products, practices, or processes. These can be explored with the very
% tools of technical interaction, such source code management systems which
% allow for exchanging not only code, but commentaries and engage in
% discussions about certain features of a particular code base. Moreover,
% these tools for technical coordination structure the interactions in
% very particular ways as they embedded assumptions about the most 
% efficient way of coordinating and combining the efforts of distributed
% software developers. Tools such as ``git'' describe the history, so
% to speak, of the development history and the developmental processes
% of the Linux kernel development, which has become a template of sorts
% for other Open Source development projects, even though the assumptions
% about gatekeeping and distibution of responsibility are not necessarily
% shared, but can be enforced through the source code management system.


% More could be said here about the intersection of "epistemological
% encounter" and "the literature" or the documentary forms at stake
% here.  In particular, the relationship to things like bug trackers,
% versioning tools repositories, project management software, chat
% software, tools fo working collaboratively etc. are "informants" in
% this sense even though they are not persons. 


\section*{Conclusion}

In this article we have discussed the particularities of fieldwork
among computer technologists and technologies with an emphasis on what
can be said in respect to the challenges and promises of field
research. The need for guidelines and specifiable rules of method for
ethnographic research tends is not always welcome in the discipline of
anthropology.   And this is the case for good reasons: one cannot
anticipate but only prepare oneself to a certain degree, to what the
field will offer in terms of shared learning, revelation and
dialog. The domain of computing is not special in this regard. What is
particular in the context of computing is the technological means
through which relations are sustained and for which technologists
dedicate their waking hours (and quite often lose their sleep). As we
presented in the previous sub-sections, these particularities have to
do with power dynamics in the field (involving computer hackers of
particular ethnic groups and ethnic and gender minorities), the
question of socioeconomic difference and prestige involving human
science researchers and elite technologists, and the dispute over the
legitimacy to represent collectives of digital experts. Other than the difficult
challenges to the ethnographic practice, there are great opportunities
for helping to advance anthropological research toward new forms of
collaborative theorizing and technical engagement---beyond more
traditional forms of collaboration in the discipline (as described by
\cite{Lassiter200x}, involving co-authoring of ethnographic texts or mere
editing of interview protocols), but advancing toward the
collaborative construction of digital infrastructures to further
collaborative research among ourselves and the constituents of our
projects (eg \cite{FortunandFortun2015}).

\bibliography{computer-experts}
\bibliographystyle{plain}

\end{document}
